---
title: "Code for CALERIE manuscript"
author: "Akshay Bareja"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggfortify)
library(biomaRt)
library(limma)
library(edgeR)
library(DT)
library(fgsea)
library(patchwork)
library(GSVA)
library(pheatmap)
library(pathview)
library(clusterProfiler)
library(org.Hs.eg.db)
library(openxlsx)
library(ggpmisc)
library(car)
```

This document contains code for the manuscript "Biological Mechanisms of Strength Preservation During Calorie Restriction-Induced Weight Loss Among Young- to Middle-Aged Adults without Obesity"

## Check renv status

```{r}
renv::status()
```

## strength ~ mass code

```{r warning = FALSE, message = FALSE}
# Import and clean data
sex_info <- read_csv("Regression Analysis Data/data/IVRSRAND.CSV") %>%   
              dplyr::select(deidnum, strata)
strength_data <- read_csv("Regression Analysis Data/data/ISOMETRA.CSV") %>% dplyr::select(deidnum, visit, ke60ptm) %>% 
              dplyr::mutate(across(where(is.character), ~ str_trim(tolower(.)))) %>%               dplyr::mutate(ke60ptm = as.numeric(ke60ptm))
mass_data <- read_csv("Regression Analysis Data/data/DXAA.CSV") %>% 
                  dplyr::select(deidnum, visit, BTOTMASS) %>% 
                  dplyr::mutate(across(where(is.character), ~ str_trim(tolower(.)))) %>% 
                  dplyr::mutate(BTOTMASS = as.numeric(BTOTMASS))

# Extract sex from 'strata' column and keep only unique `deidnum`
sex_info <- sex_info %>%
  dplyr::mutate(sex = if_else(str_detect(strata, 'Male'), 'Male', 'Female')) %>%
  dplyr::select(deidnum, sex) %>%
  dplyr::distinct(deidnum, .keep_all = TRUE)

# Filter for Baseline and Month 12 visits
baseline <- strength_data %>%
  dplyr::filter(visit == "bl sub 2") %>%
  dplyr::inner_join(dplyr::filter(mass_data, visit == "bl sub 2"), by = c('deidnum', 'visit')) 

month12 <- strength_data %>%
  dplyr::filter(visit == "month 12") %>%
  dplyr::inner_join(dplyr::filter(mass_data, visit == "month 12"), by = c('deidnum', 'visit')) 

# Rename columns for consistency
baseline <- baseline %>%
  dplyr::rename(strength_baseline = ke60ptm, mass_baseline = BTOTMASS)
month12 <- month12 %>%
  dplyr::rename(strength_12m = ke60ptm, mass_12m = BTOTMASS)

# Merge Baseline and Month 12 data
df <- baseline %>%
  dplyr::select(deidnum, strength_baseline, mass_baseline) %>%
  dplyr::inner_join(dplyr::select(month12, deidnum, strength_12m, mass_12m), 
                    join_by(deidnum))

# Merge sex info
df <- df %>%
  dplyr::left_join(sex_info, by = 'deidnum')

# Compute deltas and % changes
df <- df %>%
  dplyr::mutate(
    `∆ Strength` = strength_12m - strength_baseline,
    `% Strength` = (`∆ Strength` / strength_baseline) * 100,
    `∆ Mass` = mass_12m - mass_baseline,
    `% Mass` = (`∆ Mass` / mass_baseline) * 100
  )

# Remove unnecessary columns
df <- df %>%
  dplyr::select(deidnum, strength_baseline, strength_12m, mass_baseline, mass_12m,
         `∆ Strength`, `% Strength`, `∆ Mass`, `% Mass`, sex)

# Drop NAs for clean regression analysis
df <- df %>%
  tidyr::drop_na(`% Strength`, `% Mass`, `∆ Strength`, `∆ Mass`)

# Run regression
strength_v_mass_lm <- lm(`% Strength` ~ `% Mass`, data = df)
summary(strength_v_mass_lm)

# regression diagnostics plots
#autoplot(strength_v_mass_lm)

# R-squared
summary(strength_v_mass_lm)$r.squared

residuals_strength_v_mass <- residuals(strength_v_mass_lm) %>% 
  tibble::enframe(
    name = "deidnum", # just a placeholder here!
    value = "residuals"
  ) %>% 
  dplyr::mutate(deidnum = df$deidnum)

residuals_strength_v_mass %>% 
  dplyr::count(residuals >= 0)
# `residuals >= 0`     n
#   <lgl>            <int>
# 1 FALSE               102
# 2 TRUE                96

# save df used for regression
#write_csv(df, "results/df_for_lm_AB_012626.csv")
```

## RNAseq analysis code

### Import and clean data

```{r warning = FALSE, message = FALSE}
counts <- read_csv("Regression Analysis Data/counts_hg38.csv") %>% 
            dplyr::rename(ensembl_gene_id = `...1`) %>% 
            dplyr::rename_with(.cols = -ensembl_gene_id,
                               .fn = ~str_extract(.x, "(?<=\\_).+(?=\\_sorted)")) %>%  # clean up column names
            dplyr::rename_with(.cols = contains("month"),
                               .fn = ~str_replace(.x, "\\_month", "month")) %>% # remove underscore before each occurrence of "month"
            dplyr::rename_with(.cols = everything(),
                               .fn = ~str_replace(.x, "\\_$", "")) # needed because some sample have two underscores after the timepoint

# reshape and clean counts df
counts_long <- counts %>% 
  tidyr::pivot_longer(cols = -ensembl_gene_id,
                      names_to = "sample",
                      values_to = "count") %>% 
  dplyr::filter(!str_detect(sample, "24month")) %>% 
  tidyr::separate_wider_delim(sample, delim = "_", names = c("deidnum", "timepoint")) %>% 
  dplyr::filter(deidnum %in% residuals_strength_v_mass$deidnum) # only keep samples that are also in residual dfs

# do all subjects have two timepoints?
pairs_vec <- counts_long %>%
  dplyr::filter(ensembl_gene_id == "ENSG00000284662") %>% # picking a random gene
  dplyr::count(deidnum) %>%
  dplyr::filter(n == 2) %>%
  dplyr::pull(deidnum) # pairs
# no
# Only 42 individuals have data at both time-points

# only keep samples with counts for both timepoints (Baseline and 12M)
counts_long <- counts_long %>% 
                dplyr::filter(deidnum %in% pairs_vec)

# check if all pairs
# counts_long %>%
#   dplyr::filter(ensembl_gene_id == "ENSG00000284662") %>%
#   dplyr::count(deidnum) %>%
#   dplyr::filter(n != 2)
# yes

# look up gene symbols for Ensembl gene ids
# mart <- useEnsembl(biomart = "ensembl", 
#                    dataset = "hsapiens_gene_ensembl")
# 
# genes_df <- getBM(filters = "ensembl_gene_id",
#                   attributes = c("ensembl_gene_id", "external_gene_name"),
#                   values = counts$ensembl_gene_id, 
#                   mart = mart)
# Note - this is Ensembl v114

# save this df
#write_csv(genes_df, "metadata/genes_df_081925.csv") #v114

# import genes_df
genes_df <- read_csv("metadata/genes_df_081925.csv")

# look up gene symbols for Ensembl gene ids for Ensembl v104
# version used in Das 2023 (https://onlinelibrary.wiley.com/doi/10.1111/acel.13963)
# mart <- useEnsembl(biomart = "ensembl",
#                    dataset = "hsapiens_gene_ensembl",
#                    version = 104)
# 
# genes_df_v104 <- getBM(filters = "ensembl_gene_id",
#                   attributes = c("ensembl_gene_id", "external_gene_name"),
#                   values = counts$ensembl_gene_id,
#                   mart = mart)
# decision - use more recent annotation

# are there repeats?
# genes_df %>%
#   dplyr::count(external_gene_name) %>%
#   dplyr::arrange(desc(n)) # yes!
# genes_df %>%
#   dplyr::count(ensembl_gene_id) %>%
#   dplyr::arrange(desc(n)) # all unique

# remove genes that have 0 counts in >50% of samples
# for each subgroup i.e. baseline or 12M
# (following approach described in https://onlinelibrary.wiley.com/doi/10.1111/acel.13963)
counts_filt_df <- counts_long %>% 
  tidyr::unite(col = "sample", deidnum:timepoint, sep = "_") %>% 
  tidyr::pivot_wider(
    names_from = sample,
    values_from = count
  ) %>% 
  dplyr::rowwise() %>% 
  dplyr::mutate(zero_sums_base = sum(c_across(contains("Baseline")) == 0),
                num_samples_base = (ncol(.) - 1)/2,
                zero_frac_base = zero_sums_base/num_samples_base,
                zero_sums_12m = sum(c_across(contains("12month")) == 0),
                num_samples_12m = (ncol(.) - 1)/2,
                zero_frac_12m = zero_sums_12m/num_samples_12m) %>% 
  dplyr::filter(if_all(
    .cols = contains("frac"),
    .fns = ~.x < 0.5 # only keep genes with <50% 0 counts for BOTH timepoints
  )) %>% 
  dplyr::select(-c(matches("(_base|_12m)$"))) %>% 
  dplyr::ungroup()

# normalize counts by TMM
counts_mat <- as.matrix(counts_filt_df[ , -1])
rownames(counts_mat) <- counts_filt_df$ensembl_gene_id

## as DGEList
dge <- edgeR::DGEList(counts = counts_mat)

## calculate norm. factors
dge <- edgeR::calcNormFactors(dge,
                       method = "TMM")

## get normalized counts
counts_log_cpm <- dplyr::bind_cols(counts_filt_df[ , 1],
                                   edgeR::cpm(dge,
                                              log = TRUE))

# % of positive and negative residual individuals ALSO present in the counts dataframe
counts_log_cpm %>% 
    dplyr::filter(ensembl_gene_id == "ENSG00000005206") %>%
    tidyr::pivot_longer(cols = -ensembl_gene_id,
                        names_to = "sample",
                        values_to = "norm_count") %>% 
    tidyr::separate_wider_delim(sample, delim = "_", names = c("deidnum", "timepoint")) %>% 
    tidyr::pivot_wider(
      names_from = timepoint,
      values_from = norm_count
    ) %>% 
    dplyr::mutate(delta_norm_counts = `12month` - Baseline) %>% 
    dplyr::mutate(deidnum = as.numeric(deidnum)) %>% 
    dplyr::inner_join(residuals_strength_v_mass, 
                      join_by(deidnum)) %>% 
    dplyr::count(residuals >= 0)
# `residuals >= 0`     n
#   <lgl>            <int>
# 1 FALSE               20
# 2 TRUE                22

# save this
#write_csv(counts_log_cpm, "results/counts_log_cpm_AB_012626.csv")
```

### Fit gene expression linear models

```{r warning = FALSE, message = FALSE, include = FALSE}
# test on one gene
test_df <- counts_log_cpm %>% 
    dplyr::filter(ensembl_gene_id == "ENSG00000080824") %>% # HSP90AA1
    tidyr::pivot_longer(cols = -ensembl_gene_id,
                        names_to = "sample",
                        values_to = "norm_count") %>% 
    tidyr::separate_wider_delim(sample, delim = "_", names = c("deidnum", "timepoint")) %>% 
    tidyr::pivot_wider(
      names_from = timepoint,
      values_from = norm_count
    ) %>% 
    dplyr::mutate(delta_norm_counts = `12month` - Baseline) %>% 
    dplyr::mutate(deidnum = as.numeric(deidnum)) %>% 
    dplyr::inner_join(df, join_by(deidnum)) %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% # only keep gene IDs with corresponding gene symbols
    dplyr::filter(external_gene_name != "") 

# apply(test_df, 2, function(x) sum(is.na(x)))
# no missing values

test_lm <- lm(`% Strength` ~ delta_norm_counts + `% Mass`, data = test_df)
summary(test_lm)

# alternatively
test_lm_ancova <- lm(strength_12m ~ `12month` + strength_baseline + mass_12m + mass_baseline + Baseline, data = test_df)
summary(test_lm_ancova)

plot(test_lm)
plot(test_lm_ancova)

# pull out coefficients
summary(test_lm)$coefficients[ , 1][[2]]
summary(test_lm_ancova)$coefficients[ , 1][[2]]
```


```{r warning = FALSE, message = FALSE}
# specify model formula
formula <- reformulate(c("delta_norm_counts", "`% Mass`"), response = "`% Strength`")

# function for fitting linear models
# strength_data - data frame with strength and additional covariate data (e.g. mass, FFM...)
# counts_df - normalized gene expression dataframe 
# model_formula - model formula for lm()
fit_lm_counts_multi <- function(strength_data, counts_df, model_formula){
  
  # compute delta normalized counts
  lm_df <- counts_df %>% 
    tidyr::pivot_longer(cols = -ensembl_gene_id,
                        names_to = "sample",
                        values_to = "norm_count") %>% 
    tidyr::separate_wider_delim(sample, delim = "_", names = c("deidnum", "timepoint")) %>% 
    tidyr::pivot_wider(
      names_from = timepoint,
      values_from = norm_count
    ) %>% 
    dplyr::mutate(delta_norm_counts = `12month` - Baseline) %>% 
    dplyr::mutate(deidnum = as.numeric(deidnum)) %>% 
    dplyr::inner_join(strength_data, join_by(deidnum)) %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% # only keep gene IDs with corresponding gene symbols
    dplyr::filter(external_gene_name != "") %>% 
    dplyr::nest_by(ensembl_gene_id)
  
  # fit linear model for all genes
  models_df <- lm_df %>% 
    dplyr::mutate(mod = list(lm(model_formula, data = data)))
  
  # pull out results
  results_model_df <- models_df %>% 
    summarise(beta = summary(mod)$coefficients[ , 1][[2]], # pull out coefficient corresponding to delta_norm_counts
              std_error = summary(mod)$coefficients[ , 2][[2]],
              t_stat = summary(mod)$coefficients[ , 3][[2]],
              p_val = summary(mod)$coefficients[ , 4][[2]]) %>% 
                      dplyr::ungroup()
  
  # add BH-corrected p-values
  results_model_df$BH_p_val <- p.adjust(results_model_df$p_val,
                                        method = "BH")
  
  return(results_model_df)
  
}

results_gene_mass <- fit_lm_counts_multi(strength_data = df,
                                         counts_df = counts_log_cpm,
                                         model_formula = formula)

# present results
results_gene_mass %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::arrange(p_val) %>% 
    dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
    DT::datatable(caption = "% Strength ~ delta_norm_counts results (CPM) + % Mass")

# number of results with p < 0.01
results_gene_mass %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::filter(p_val < 0.01)
# 151

# N highest positive t-stats and highest negative t-stats
results_gene_mass %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>%
  dplyr::group_by(t_stat > 0) %>% 
  dplyr::slice_max(abs(t_stat), n = 10)

# save results
# write_csv(results_gene_mass, "results/results_gene_mass_AB_012626.csv")
```

### GSEA

```{r warning = FALSE, message = FALSE}
# gmt files downloaded from -
# https://data.broadinstitute.org/gsea-msigdb/msigdb/release/2024.1.Hs/

# write function for running fgsea
gsea_fun <- function(lm_results_df, ranking_metric, gene_set){
  
  if(ranking_metric == "t-stat"){
    
    # ranking metric - average t-stat for each gene symbol
    ranks <- lm_results_df %>% 
      dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
      dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
      dplyr::group_by(external_gene_name) %>% 
      dplyr::summarise(mean_t = mean(t_stat)) %>% 
      dplyr::arrange(desc(mean_t)) %>% 
      tibble::deframe()
    
  } else if (ranking_metric == "sbp"){
    
    # ranking metric - average signed fold change * -log10(p_val)
    ranks <- lm_results_df %>% 
      dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
      dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>%
      dplyr::mutate(sign_beta_pval = sign(beta)*-log10(p_val)) %>% 
      dplyr::group_by(external_gene_name) %>% 
      dplyr::summarise(mean_sbp = mean(sign_beta_pval)) %>% 
      dplyr::arrange(desc(mean_sbp)) %>% 
      tibble::deframe()
    
  }
   
  if(gene_set == "Hallmark"){
      # import pathway gmt files
      # Hallmark
      pathways.h.all.v2024 <- gmtPathways("metadata/h.all.v2024.1.Hs.symbols.gmt.txt")
    
      set.seed(42)
      fgseaRes_hallmark <- fgseaMultilevel(pathways = pathways.h.all.v2024, 
                                           stats = ranks) %>% 
                           dplyr::arrange(pval)
      
      return(fgseaRes_hallmark)
      
  } else if(gene_set == "Everything"){
    
      # Everything
      pathways.msigdb.v2024 <- gmtPathways("metadata/msigdb.v2024.1.Hs.symbols.gmt.txt")
      
      set.seed(43)
      fgseaRes_msigdb <- fgseaMultilevel(pathways = pathways.msigdb.v2024, 
                                         stats = ranks) %>% 
                         dplyr::arrange(pval)
    
      return(fgseaRes_msigdb)
      
  }
  
}

# Hallmark results
results_gene_mass_gsea_tstat <- gsea_fun(lm_results_df = results_gene_mass, 
                                             ranking_metric = "t-stat", 
                                             gene_set = "Hallmark") 

results_gene_mass_gsea_tstat %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  DT::datatable(caption = "SVM - Hallmark, CPM, t-stat ranks")

# save results
# saveRDS(results_gene_mass_gsea_tstat, "results/results_gene_mass_gsea_tstat_gseahallmark_012626.rds")
```

### GSVA

Perform subject-level pathway analysis and note pathways that change dramatically for high-residual individuals.

Subjects to prioritise - 
High positive-residual individuals - 
94495
68089
44288

High negative-residual individuals - 
47346
79877
50384

```{r warning = FALSE, message = FALSE}
# Prep data for GSVA
# convert Ensembl IDs to gene symbols
# where you have multiple gene symbols mapping to the same ensembl ID, keep the one with highest variance across all samples
counts_log_cpm_gsva <- counts_log_cpm %>% 
                    dplyr::inner_join(
                      genes_df,
                      join_by(ensembl_gene_id)
                    ) %>% 
                    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
                    dplyr::select(-ensembl_gene_id) %>% 
                    dplyr::rowwise() %>% 
                    dplyr::mutate(gene_var = var(c_across(where(is.numeric)), na.rm = TRUE)) %>% 
                    dplyr::ungroup() %>% 
                    dplyr::group_by(external_gene_name) %>% 
                    dplyr::slice_max(gene_var) %>% 
                    dplyr::ungroup() %>% 
                    dplyr::select(-gene_var) %>% 
                    tidyr::drop_na()

# Note: we go from 32291 to 24867 genes when we pick the duplicate with highest variance

counts_lcpm_mat <- as.matrix(counts_log_cpm_gsva[-1])
rownames(counts_lcpm_mat) <- counts_log_cpm_gsva$external_gene_name

# get gene sets
pathways.h.all.v2024 <- gmtPathways("metadata/h.all.v2024.1.Hs.symbols.gmt.txt")
pathways.msigdb.v2024 <- gmtPathways("metadata/msigdb.v2024.1.Hs.symbols.gmt.txt")

# HALLMARK
gsva_hallmark <- GSVA::gsvaParam(exprData = counts_lcpm_mat,
                                 geneSets = pathways.h.all.v2024,
                                 minSize = 5,
                                 kcdf = "Gaussian")
gsva_hallmark_res <- GSVA::gsva(gsva_hallmark)

# inspect results
gsva_hallmark_res_reshape <- gsva_hallmark_res %>% 
  data.frame() %>% 
  tibble::rownames_to_column("pathway") %>% 
  tidyr::pivot_longer(
    cols = -pathway,
    names_to = "id_time",
    values_to = "score"
  ) %>% 
  dplyr::mutate(id_time = str_remove(id_time, "^X")) %>% 
  tidyr::separate_wider_delim(
    cols = id_time,
    names = c("deidnum", "timepoint"),
    delim = "_"
  ) %>% 
  tidyr::pivot_wider(
    names_from = timepoint,
    values_from = score
  ) %>% 
  dplyr::mutate(score_delta = `12month` - Baseline)

# compare extreme positive v negative residual individuals
# positive
gsva_hallmark_res_reshape %>% 
  dplyr::filter(deidnum == 44288) %>% View()
# negative
gsva_hallmark_res_reshape %>% 
  dplyr::filter(deidnum ==  50384) %>% View()

# Heatmap of GSVA scores for extreme-residual individuals

# prep heatmap mat
heatmap_df <- gsva_hallmark_res_reshape %>% 
  dplyr::filter(deidnum %in% c(94495, 68089, 44288, 47346, 79877, 50384)) %>% 
  dplyr::select(pathway, deidnum, score_delta) %>% 
  tidyr::pivot_wider(
    names_from = deidnum,
    values_from = score_delta
  )

heatmap_mat <- as.matrix(heatmap_df[-1])
rownames(heatmap_mat) <- heatmap_df$pathway

# make column annotation dd
annot_col <- tibble::tibble(
  residual = c("pos", "neg", "pos", "pos", "neg", "neg")
) %>% 
  data.frame()
rownames(annot_col) <- colnames(heatmap_mat)

pheatmap(heatmap_mat,
         annotation_col = annot_col)

# Correlating GSVA score deltas with residuals
# reshape GSVA df
gsva_cor_df <- gsva_hallmark_res_reshape %>% 
                dplyr::select(pathway, deidnum, score_delta) %>% 
                tidyr::pivot_wider(
                  names_from = pathway,
                  values_from = score_delta
                ) %>% 
                dplyr::mutate(deidnum = as.numeric(deidnum)) %>% 
                dplyr::arrange(deidnum)

# keep same deidnum and order in residuals df
residuals_cor_df <- residuals_strength_v_mass %>% 
                      dplyr::filter(deidnum %in% gsva_cor_df$deidnum) %>% 
                      dplyr::arrange(deidnum)

# ensure equal deidnum order
# identical(residuals_cor_df$deidnum, gsva_cor_df$deidnum) # TRUE

# Pearson correlation coefficients
cor(residuals_cor_df$residuals, gsva_cor_df[-1]) %>% 
  data.frame() %>% 
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "pathway",
    values_to = "cor_coeff"
  ) %>% View()
```

## Top ten protein-coding genes

i.e. top 5 positive and top 5 negative protein-coding genes

```{r message = FALSE, warning = FALSE}
# mart <- useEnsembl(biomart = "ensembl",
#                    dataset = "hsapiens_gene_ensembl",
#                    version = 114)

# Get biotypes
# gene_biotypes <- biomaRt::getBM(
#   attributes = c("ensembl_gene_id", "gene_biotype"),
#   filters = "ensembl_gene_id",
#   values = unique(results_gene_mass$ensembl_gene_id),
#   mart = mart
# )

# save this file
# write_csv(gene_biotypes, "metadata/gene_biotypes_012626.csv")

# import gene biotypes
gene_biotypes <- read_csv("metadata/gene_biotypes_012626.csv")

# Keep only protein-coding genes
protein_coding_ids <- gene_biotypes %>%
  dplyr::filter(gene_biotype == "protein_coding") %>%
  dplyr::pull(ensembl_gene_id)

# get top 5 positive protein-coding genes
top_genes_cpm_protein <- results_gene_mass %>%
  dplyr::ungroup() %>%
  dplyr::filter(
    ensembl_gene_id %in% protein_coding_ids,
    p_val < 0.01
  ) %>%
  dplyr::arrange(desc(t_stat)) %>%
  dplyr::slice_head(n = 5)

# get top 5 negative protein-coding genes
bottom_genes_cpm_protein <- results_gene_mass %>%
  dplyr::ungroup() %>%
  dplyr::filter(
    ensembl_gene_id %in% protein_coding_ids,
    p_val < 0.01
  ) %>%
  dplyr::arrange(t_stat) %>%
  dplyr::slice_head(n = 5)

# combine positive and negative PC genes into one table
top10_genes_cpm_protein <- dplyr::bind_rows(
  top_genes_cpm_protein, bottom_genes_cpm_protein) %>%
  dplyr::left_join(genes_df, 
                   join_by(ensembl_gene_id)) %>%
  dplyr::relocate(external_gene_name, .after = ensembl_gene_id)

# save this
# write_csv(top10_genes_cpm_protein, "results/top10_genes_cpm_protein_NM_012626.csv")

# This is the file NM used for her STRING analysis
```

## Some values reported in the paper

```{r message = FALSE, warning = FALSE}
# total number of starting genes
nrow(counts) # 60,605

# total number of filtered genes
nrow(counts_log_cpm) # 32,291

# total number of filtered genes with corresponding gene symbols
nrow(results_gene_mass) # 25,200
```

## Figures and tables for manuscript

```{r message = FALSE, warning = FALSE}
# import data and results
df <- read_csv("results/df_for_lm_AB_012626.csv")
SUBJECT1 <- read_csv("Regression Analysis Data/data/SUBJECT1.CSV") %>%  # to get age
  dplyr::select(deidnum, age)
CLWTVIS <- read_csv("Regression Analysis Data/data/CLWTVIS.CSV") %>%  # to get BMI (in `mbmi` column)
  dplyr::filter(visit %in% c("Baseline", "Month 12")) %>%
  dplyr::select(deidnum, visit, mbmi) %>% 
  tidyr::pivot_wider(
    names_from = visit,
    values_from = mbmi
  )
counts_log_cpm <- read_csv("results/counts_log_cpm_AB_012626.csv")
results_gene_mass <- read_csv("results/results_gene_mass_AB_012626.csv")
genes_df <- read_csv("metadata/genes_df_081925.csv")
results_gene_mass_gsea_tstat <- readRDS("results/results_gene_mass_gsea_tstat_gseahallmark_012626.rds")
top10_genes_cpm_protein <- read_csv("results/top10_genes_cpm_protein_NM_012626.csv")

# Table 1
## Strength vs. Whole-Body Mass Cohort (N = 198)
df %>% 
  dplyr::inner_join(
    SUBJECT1,
    join_by(deidnum)
  ) %>% 
  dplyr::inner_join(
    CLWTVIS,
    join_by(deidnum)
  ) %>% 
  dplyr::summarise(
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    mean_bmi_base = mean(Baseline, na.rm = TRUE),
    sd_bmi_base = sd(Baseline, na.rm = TRUE),
    mean_bmi_12m = mean(`Month 12`, na.rm = TRUE),
    sd_bmi_12m = sd(`Month 12`, na.rm = TRUE),
    mean_perc_strength_change = mean(`% Strength`),
    sd_perc_strength_change = sd(`% Strength`),
    mean_perc_wbm_change = mean(`% Mass`),
    sd_perc_wbm_change = sd(`% Mass`)
  ) %>% View()

df %>% 
  dplyr::count(sex) %>% 
  dplyr::mutate(perc = 100 * n/sum(n) )
  
## Residuals vs. Gene Transcripts Cohort (N = 42)
transcript_deidnums <- as.numeric(unique(str_extract(colnames(counts_log_cpm)[-1], ".+(?=\\_)")))
                        
df %>% 
  dplyr::inner_join(
    SUBJECT1,
    join_by(deidnum)
  ) %>% 
  dplyr::inner_join(
    CLWTVIS,
    join_by(deidnum)
  ) %>% 
  dplyr::filter(deidnum %in% transcript_deidnums) %>% 
  dplyr::summarise(
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    mean_bmi_base = mean(Baseline, na.rm = TRUE),
    sd_bmi_base = sd(Baseline, na.rm = TRUE),
    mean_bmi_12m = mean(`Month 12`, na.rm = TRUE),
    sd_bmi_12m = sd(`Month 12`, na.rm = TRUE),
    mean_perc_strength_change = mean(`% Strength`),
    sd_perc_strength_change = sd(`% Strength`),
    mean_perc_wbm_change = mean(`% Mass`),
    sd_perc_wbm_change = sd(`% Mass`)
  ) %>% View()
 
df %>% 
  dplyr::filter(deidnum %in% transcript_deidnums) %>%
  dplyr::count(sex) %>% 
  dplyr::mutate(perc = 100 * n/sum(n) ) 

# Table 2
# top 5 positive and negative NES pathways
table_2 <- results_gene_mass_gsea_tstat %>% 
  dplyr::group_by(NES > 0) %>% 
  dplyr::slice_min(pval, n = 5) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(-leadingEdge) %>%
  dplyr::mutate(across(.cols = -c(pathway:padj),
                       .fns = ~round(.x, 2))) %>% 
  dplyr::mutate(
    pathway = str_remove(pathway, "HALLMARK_"),
    pathway = str_replace_all(pathway, "_", " ")
  ) %>% 
  dplyr::rename(
    `Hallmark Pathway` = pathway,
    `p-value` = pval,
    `Adjusted p-value` = padj,
    `Log2 Error` = log2err,
    Size = size
  ) %>% 
  dplyr::select(`Hallmark Pathway`:Size) 
# save table 2
# write.xlsx(table_2, "results/table_2_AB_012626.xlsx")

# Table 3
table_3 <- top10_genes_cpm_protein %>% 
  dplyr::select(external_gene_name:p_val) %>% 
  dplyr::mutate(across(.cols = beta:t_stat,
                       .fns = ~ round(.x, 2))) %>% 
  dplyr::rename(
    `Gene Name` = external_gene_name,
    Beta = beta,
    `Standard Error` = std_error,
    `t-statistic` = t_stat,
    `p-value` = p_val
  )
# save table 3
# write.xlsx(table_3, "results/table_3_AB_012626.xlsx")

# Figure 1
# "results/figures/figure_1_flowchart_for_CALERIE.001.jpegg"
# Made by AB

# Figure 2
figure_2 <- df %>% 
  dplyr::mutate(
    group_omics = if_else(
      deidnum %in% transcript_deidnums,
      "Has transcriptomics data",
      "Does not have transcriptomics data"
    )
  ) %>% 
  ggplot(aes(`% Mass`, `% Strength`)) +
  geom_point(aes(col = group_omics)) +
  geom_smooth(method = "lm",
              se = FALSE,
              linetype = "dashed",
              col = "black") +
  geom_vline(xintercept = 0,
             linetype = "dotted") +
  geom_hline(yintercept = 0,
             linetype = "dotted") +
  theme_bw() +
  labs(x = "% Change in Whole-Body Mass",
       y = "% Change in Strength") +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  stat_poly_eq(aes(label = after_stat(rr.label)), formula = y ~ x, parse = TRUE)

# save Figure 2
# ggsave("results/figures/figure_2_AB_012626.tiff", figure_2)

# Figure 3
figure_3 <- residuals_strength_v_mass %>% 
  ggplot(aes(residuals)) +
  geom_histogram(aes(y = after_stat(count)),
                 col = "black",
                 fill = "lightblue") +
  geom_vline(xintercept = mean(residuals_strength_v_mass$residuals),
             linetype = "dashed") +
  theme_bw() +
  labs(x = "Residual Value",
       y = "Count")
# save Figure 3
# ggsave("results/figures/figure_3_AB_012626.tiff", figure_3)

# Figure 4
figure_4 <- table_3 %>% 
  dplyr::mutate(direction = if_else(NES > 0,
                                    "Upregulated", "Downregulated"),
                direction = fct_relevel(direction,
                                        "Upregulated")) %>% 
  ggplot(aes(NES, fct_reorder(`Hallmark Pathway`, NES),
             fill = direction)) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(values = c("darkgreen", "darkred")) +
  labs(x = "Normalized Enriched Score (NES)",
       y = "") +
  theme(legend.title = element_blank())
# save Figure 4
# ggsave("results/figures/figure_4_AB_012626.tiff", figure_4)

# Figure 5
# "results/figures/string_result_1_21_26.png"
```

## Additional analyses

### Adding sex as a covariate

```{r message = FALSE, warning = FALSE}
# fit gene expression linear models
# specify model formula
formula_sex <- reformulate(c("delta_norm_counts", "`% Mass`", "sex"), response = "`% Strength`")

results_gene_mass_sex <- fit_lm_counts_multi(strength_data = df,
                                             counts_df = counts_log_cpm,
                                             model_formula = formula_sex)

# present results
results_gene_mass_sex %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::arrange(p_val) %>% 
    dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
    DT::datatable(caption = "% Strength ~ delta_norm_counts results (CPM) + % Mass + sex")

# number of results with p < 0.01
results_gene_mass_sex %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::filter(p_val < 0.01)
# 143

# N highest positive t-stats and highest negative t-stats
results_gene_mass_sex %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>%
  dplyr::group_by(t_stat > 0) %>% 
  dplyr::slice_max(abs(t_stat), n = 5)

# compare with % Mass data
results_gene_mass %>% 
  dplyr::inner_join(
    results_gene_mass_sex,
    join_by(ensembl_gene_id),
    suffix = c("_mass", "_mass_sex")
  ) %>% 
  ggplot(aes(beta_mass, beta_mass_sex)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Beta estimates for % Mass vs % Mass + sex")
```

### % strength change ~ % FFM change

```{r message = FALSE, warning = FALSE}
ffm_data <- read_csv("Regression Analysis Data/data/DXAA.CSV") %>% 
              dplyr::select(deidnum, visit, BTOTFFM) %>% 
              dplyr::mutate(across(where(is.character), ~ str_trim(tolower(.)))) %>% 
              dplyr::mutate(BTOTFFM = as.numeric(BTOTFFM)) %>% 
              dplyr::filter(visit %in% c("bl sub 2", "month 12")) %>% 
              tidyr::pivot_wider(
                names_from = "visit",
                values_from = "BTOTFFM"
              ) %>% 
              tidyr::drop_na() %>% 
              dplyr::rename(
                ffm_baseline = `bl sub 2`,
                ffm_12m = `month 12`
              ) %>% 
              dplyr::mutate(
                  `∆ FFM` = ffm_12m - ffm_baseline,
                  `% FFM` = (`∆ FFM` / ffm_baseline) * 100
                )

# join with main df
df_with_ffm <- df %>% 
                dplyr::left_join(
                  ffm_data,
                  join_by(deidnum)
                )

# Run regression
strength_v_ffm_lm <- lm(`% Strength` ~ `% FFM`, data = df_with_ffm)
summary(strength_v_ffm_lm) # much higher coefficient compared to % Mass model

# fit gene expression linear models
# specify model formula
formula_ffm <- reformulate(c("delta_norm_counts", "`% FFM`"), response = "`% Strength`")

results_gene_ffm <- fit_lm_counts_multi(strength_data = df_with_ffm,
                                         counts_df = counts_log_cpm,
                                         model_formula = formula_ffm)

# present results
results_gene_ffm %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::arrange(p_val) %>% 
    dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
    DT::datatable(caption = "% Strength ~ delta_norm_counts results (CPM) + % FFM")

# number of results with p < 0.01
results_gene_ffm %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::filter(p_val < 0.01)
# 169

# N highest positive t-stats and highest negative t-stats
results_gene_ffm %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>%
  dplyr::group_by(t_stat > 0) %>% 
  dplyr::slice_max(abs(t_stat), n = 5)

# compare with % Mass data
mass_v_ffm_fig <- results_gene_mass %>% 
  dplyr::inner_join(
    results_gene_ffm,
    join_by(ensembl_gene_id),
    suffix = c("_mass", "_ffm")
  ) %>% 
  ggplot(aes(beta_mass, beta_ffm)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Beta estimates for % Mass vs % FFM")
```

### % strength change ~ % LSTM change

lean soft-tissue mass = BTOTFFM - BTOTBMC

```{r message = FALSE, warning = FALSE}
bmc_data <- read_csv("Regression Analysis Data/data/DXAA.CSV") %>% 
              dplyr::select(deidnum, visit, BTOTBMC) %>% 
              dplyr::mutate(across(where(is.character), ~ str_trim(tolower(.)))) %>% 
              dplyr::mutate(BTOTBMC = as.numeric(BTOTBMC)) %>% 
              dplyr::filter(visit %in% c("bl sub 2", "month 12")) %>% 
              tidyr::pivot_wider(
                names_from = "visit",
                values_from = "BTOTBMC"
              ) %>% 
              tidyr::drop_na() %>% 
              dplyr::rename(
                bmc_baseline = `bl sub 2`,
                bmc_12m = `month 12`
              )

# join with df_with_ffm and compute lstm changes
df_with_lstm <- df_with_ffm %>% 
  dplyr::left_join(
    bmc_data,
    join_by(deidnum)
  ) %>% 
  dplyr::mutate(
    `∆ LSTM` = (ffm_12m - bmc_12m) - (ffm_baseline - bmc_baseline),
    `% LSTM` = (`∆ LSTM` / (ffm_baseline - bmc_baseline)) * 100
  ) 

# Run regression
strength_v_lstm_lm <- lm(`% Strength` ~ `% LSTM`, data = df_with_lstm)
summary(strength_v_lstm_lm) # much higher coefficient compared to % Mass model

# fit gene expression linear models
# specify model formula
formula_lstm <- reformulate(c("delta_norm_counts", "`% LSTM`"), response = "`% Strength`")

results_gene_lstm <- fit_lm_counts_multi(strength_data = df_with_lstm,
                                         counts_df = counts_log_cpm,
                                         model_formula = formula_lstm)

# present results
results_gene_lstm %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::arrange(p_val) %>% 
    dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
    DT::datatable(caption = "% Strength ~ delta_norm_counts results (CPM) + % LSTM")

# number of results with p < 0.01
results_gene_lstm %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::filter(p_val < 0.01)
# 165

# N highest positive t-stats and highest negative t-stats
results_gene_lstm %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>%
  dplyr::group_by(t_stat > 0) %>% 
  dplyr::slice_max(abs(t_stat), n = 5)

# compare with % Mass data
mass_vs_lstm_fig <- results_gene_mass %>% 
  dplyr::inner_join(
    results_gene_lstm,
    join_by(ensembl_gene_id),
    suffix = c("_mass", "_lstm")
  ) %>% 
  ggplot(aes(beta_mass, beta_lstm)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Beta estimates for % Mass vs % LSTM")
```

**Conclusion:** All of the above results are very similar to those of the model presented in the paper. 

### Looking only at individuals who lost mass

```{r}
df_mass_loss <- df %>% 
                  dplyr::filter(`% Mass` < 0)
nrow(df_mass_loss) # 156

# how many of these individuals also have transcriptomics data?
counts_log_cpm %>% 
    dplyr::filter(ensembl_gene_id == "ENSG00000155657") %>%
    tidyr::pivot_longer(cols = -ensembl_gene_id,
                        names_to = "sample",
                        values_to = "norm_count") %>% 
    tidyr::separate_wider_delim(sample, delim = "_", names = c("deidnum", "timepoint")) %>% 
    tidyr::pivot_wider(
      names_from = timepoint,
      values_from = norm_count
    ) %>% 
    dplyr::mutate(delta_norm_counts = `12month` - Baseline) %>% 
    dplyr::mutate(deidnum = as.numeric(deidnum)) %>% 
    dplyr::inner_join(df_mass_loss, join_by(deidnum)) %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% # only keep gene IDs with corresponding gene symbols
    dplyr::filter(external_gene_name != "") %>% 
    nrow() 
# 33

# we lost 9 individuals

# fit gene expression linear models
# specify model formula
formula_mass <- reformulate(c("delta_norm_counts", "`% Mass`"), response = "`% Strength`")

results_gene_mass_loss <- fit_lm_counts_multi(strength_data = df_mass_loss,
                                         counts_df = counts_log_cpm,
                                         model_formula = formula_mass)

# present results
results_gene_mass_loss %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::arrange(p_val) %>% 
    dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
    DT::datatable(caption = "% Strength ~ delta_norm_counts results (CPM) + % Mass")

# number of results with p < 0.01
results_gene_mass_loss %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::filter(p_val < 0.01)
# 77

# N highest positive t-stats and highest negative t-stats
results_gene_mass_loss %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>%
  dplyr::group_by(t_stat > 0) %>% 
  dplyr::slice_max(abs(t_stat), n = 5)

# compare with % Mass data
all_vs_mass_loss_fig <- results_gene_mass %>% 
  dplyr::inner_join(
    results_gene_mass_loss,
    join_by(ensembl_gene_id),
    suffix = c("_mass", "_mass_loss")
  ) %>% 
  ggplot(aes(beta_mass, beta_mass_loss)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Beta estimates for Full vs Only Mass Loss")
```

Pretty decent agreement

### Adjusting for baseline strength

```{r message = FALSE, warning = FALSE}
# fit gene expression linear models
# specify model formula
formula_base <- reformulate(c("delta_norm_counts", "`% Mass`", "strength_baseline"), response = "`% Strength`")

results_gene_mass_base <- fit_lm_counts_multi(strength_data = df,
                                             counts_df = counts_log_cpm,
                                             model_formula = formula_base)

# present results
results_gene_mass_base %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::arrange(p_val) %>% 
    dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
    DT::datatable(caption = "% Strength ~ delta_norm_counts results (CPM) + % Mass + strength_baseline")

# number of results with p < 0.01
results_gene_mass_base %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::filter(p_val < 0.01)
# 140

# N highest positive t-stats and highest negative t-stats
results_gene_mass_base %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>%
  dplyr::group_by(t_stat > 0) %>% 
  dplyr::slice_max(abs(t_stat), n = 5)

# compare with % Mass data
mass_vs_baseline_fig <- results_gene_mass %>% 
  dplyr::inner_join(
    results_gene_mass_base,
    join_by(ensembl_gene_id),
    suffix = c("_mass", "_mass_base")
  ) %>% 
  ggplot(aes(beta_mass, beta_mass_base)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Beta estimates for % Mass vs % Mass + strength_baseline")
```

Good agreement.  

### ANCOVA-style model

Fitting the following model - 

strength_12m ~ `12month` + strength_baseline + mass_baseline + mass_12m + Baseline

Where - 
`12month` = logCPM at 12 month
Baseline = logCPM at baseline

Note - it's critical that the **first** term in the model formula above be the 12-month logCPM value!!

```{r warning = FALSE, message = FALSE}
# fit gene expression linear models
# specify model formula
formula_ancova <- reformulate(c("`12month`", "strength_baseline", "mass_baseline", "mass_12m", "Baseline"), response = "strength_12m")

results_gene_mass_ancova <- fit_lm_counts_multi(strength_data = df,
                                                counts_df = counts_log_cpm,
                                                model_formula = formula_ancova)

# present results
results_gene_mass_ancova %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::arrange(p_val) %>% 
    dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
    DT::datatable(caption = "ANCOVA model")

# number of results with p < 0.01
results_gene_mass_ancova %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::filter(p_val < 0.01)
# 106

# N highest positive t-stats and highest negative t-stats
results_gene_mass_ancova %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>%
  dplyr::group_by(t_stat > 0) %>% 
  dplyr::slice_max(abs(t_stat), n = 5)

# compare with % Mass data
results_gene_mass %>% 
  dplyr::inner_join(
    results_gene_mass_ancova,
    join_by(ensembl_gene_id),
    suffix = c("_mass", "_mass_ancova")
  ) %>% 
  ggplot(aes(beta_mass, beta_mass_ancova)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Beta estimates for % Mass vs ANCOVA model")
# ok agreement
```

#### Comparing first model with ANCOVA-style model for a few top genes

```{r message = FALSE, warning = FALSE}
# genes to test
# top most significant genes for first model
# ENSG00000080824 (HSP90AA1)
# ENSG00000158417 (EIF5B)
# ENSG00000211459 (MT-RNR1)
# ENSG00000226803 (ZNF451-AS1)
# ENSG00000210082 (MT-RNR2)

# top most significant genes for ANCOVA-style model
# ENSG00000251298 (TMEM184C-DT)
# ENSG00000268758 (ADGRE4P)
# ENSG00000234336 (JAZF1-AS1)
# ENSG00000276404 (MIR6835)
# ENSG00000115461 (IGFBP5)

# test on one gene
test_df <- counts_log_cpm %>% 
    dplyr::filter(ensembl_gene_id == "ENSG00000080824") %>% 
    tidyr::pivot_longer(cols = -ensembl_gene_id,
                        names_to = "sample",
                        values_to = "norm_count") %>% 
    tidyr::separate_wider_delim(sample, delim = "_", names = c("deidnum", "timepoint")) %>% 
    tidyr::pivot_wider(
      names_from = timepoint,
      values_from = norm_count
    ) %>% 
    dplyr::mutate(delta_norm_counts = `12month` - Baseline) %>% 
    dplyr::mutate(deidnum = as.numeric(deidnum)) %>% 
    dplyr::inner_join(df, join_by(deidnum)) %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% # only keep gene IDs with corresponding gene symbols
    dplyr::filter(external_gene_name != "") 

# apply(test_df, 2, function(x) sum(is.na(x)))
# no missing values

test_lm <- lm(`% Strength` ~ delta_norm_counts + `% Mass`, data = test_df)
summary(test_lm)

test_lm_delta <- lm(`∆ Strength` ~ delta_norm_counts + `∆ Mass`, data = test_df)
summary(test_lm_delta)

# alternatively
test_lm_ancova <- lm(strength_12m ~ `12month` + strength_baseline + mass_12m + mass_baseline + Baseline, data = test_df)
summary(test_lm_ancova)

plot(test_lm)
plot(test_lm_delta)
plot(test_lm_ancova)

# remove deidnum and re-assess
test_df_no_94495 <- test_df %>% 
                      dplyr::filter(deidnum != "94495")
test_lm_no_94495 <- lm(`% Strength` ~ delta_norm_counts + `% Mass`, data = test_df_no_94495)
test_lm_ancova_no_94495 <- lm(strength_12m ~ `12month` + strength_baseline + mass_12m + mass_baseline + Baseline, data = test_df_no_94495)

summary(test_lm_no_94495)
summary(test_lm_ancova_no_94495)

plot(test_lm_no_94495)
plot(test_lm_ancova_no_94495)

# removing this individual improves the plots a bit
```

Notes - 

The ANCOVA-style model produces better diagnostic plots, **especially** for ENSG00000080824 (HSP90AA1)!!

**Note:** deidnum 94495 looks like an influential point in a number of analyses. For example, after removing this individual, ENSG00000211459 goes from being "highly significant" to not significant at all!

**Decision: **
- Repeat ANCOVA-style model AFTER removing deidnum 94495

### ANCOVA-style model WITHOUT deidnum 94495

```{r message = FALSE, warning = FALSE}
# filter counts df
counts_df_no_94495 <- counts_log_cpm %>% 
                        dplyr::select(!contains("94495"))

# fit gene expression linear models
# specify model formula
formula_ancova <- reformulate(c("`12month`", "strength_baseline", "mass_baseline", "mass_12m", "Baseline"), response = "strength_12m")

results_gene_mass_ancova_no_94495 <- fit_lm_counts_multi(strength_data = df,
                                                counts_df = counts_df_no_94495,
                                                model_formula = formula_ancova)

# present results
results_gene_mass_ancova_no_94495 %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::arrange(p_val) %>% 
    dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>% 
    DT::datatable(caption = "ANCOVA model (no 94495)")

# number of results with p < 0.01
results_gene_mass_ancova_no_94495 %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>% 
    dplyr::filter(p_val < 0.01)
# 106

# N highest positive t-stats and highest negative t-stats
results_gene_mass_ancova_no_94495 %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% 
    dplyr::relocate(external_gene_name, .after = ensembl_gene_id) %>%
  dplyr::group_by(t_stat > 0) %>% 
  dplyr::slice_max(abs(t_stat), n = 5)

# compare with % Mass data
results_gene_mass %>% 
  dplyr::inner_join(
    results_gene_mass_ancova_no_94495,
    join_by(ensembl_gene_id),
    suffix = c("_mass", "_mass_ancova")
  ) %>% 
  ggplot(aes(beta_mass, beta_mass_ancova)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Beta estimates for % Mass vs ANCOVA model")
# ok agreement

# compare with ANCOVA-style model using full data
results_gene_mass_ancova %>% 
  dplyr::inner_join(
    results_gene_mass_ancova_no_94495,
    join_by(ensembl_gene_id),
    suffix = c("_ancova", "_ancova_no_94495")
  ) %>% 
  ggplot(aes(beta_ancova, beta_ancova_no_94495)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "Beta estimates for ANCOVA models")
# again, ok agreement
```

#### GSEA

```{r warning = FALSE, message = FALSE}
# gmt files downloaded from -
# https://data.broadinstitute.org/gsea-msigdb/msigdb/release/2024.1.Hs/

# Hallmark results
results_gene_ancova_gsea_tstat <- gsea_fun(lm_results_df = results_gene_mass_ancova_no_94495, 
                                             ranking_metric = "t-stat", 
                                             gene_set = "Hallmark") 

results_gene_ancova_gsea_tstat %>% 
  dplyr::mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  DT::datatable(caption = "ANCOVA - Hallmark, CPM, t-stat ranks")

```

### Added variables plot for HSP90AA1

ENSG00000080824

```{r message = FALSE, warning = FALSE}
# test on one gene
hsp90aa1_df <- counts_log_cpm %>% 
    dplyr::filter(ensembl_gene_id == "ENSG00000080824") %>%
    tidyr::pivot_longer(cols = -ensembl_gene_id,
                        names_to = "sample",
                        values_to = "norm_count") %>% 
    tidyr::separate_wider_delim(sample, delim = "_", names = c("deidnum", "timepoint")) %>% 
    tidyr::pivot_wider(
      names_from = timepoint,
      values_from = norm_count
    ) %>% 
    dplyr::mutate(delta_norm_counts = `12month` - Baseline) %>% 
    dplyr::mutate(deidnum = as.numeric(deidnum)) %>% 
    dplyr::inner_join(df, join_by(deidnum)) %>% 
    dplyr::inner_join(genes_df, join_by(ensembl_gene_id)) %>% # only keep gene IDs with corresponding gene symbols
    dplyr::filter(external_gene_name != "") 

hsp90aa1_lm <- lm(`% Strength` ~ delta_norm_counts + `% Mass`, data = hsp90aa1_df)
summary(hsp90aa1_lm)

avplot_data <- car::avPlot(model = hsp90aa1_lm,
                      variable = "delta_norm_counts")

avplot_data %>% 
  ggplot(aes(delta_norm_counts, `% Strength`)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE) +
  theme_bw() +
  labs(title = "Added Variable Plot for delta HSP90AA1 counts",
       x = "Residuals of delta HSP90AA1 counts regressed on % Mass",
       y = "Residuals of % Strength regressed on % Mass")
```

## Supplementary Figures and tables

```{r}
supp_fig_1 <- (mass_v_ffm_fig / mass_vs_lstm_fig / all_vs_mass_loss_fig / mass_vs_baseline_fig) +
                  plot_annotation(tag_level = "A")

# save Supplementary Figure 1
ggsave("results/figures/supp_figure_1_AB_033126.tiff",
       supp_fig_1,
       height = 9)
```


## Session Info

```{r}
utils::sessionInfo()

# R version 4.5.0 (2025-04-11)
# Platform: aarch64-apple-darwin20
# Running under: macOS Sonoma 14.4
# 
# Matrix products: default
# BLAS:   /System/Library/Frameworks/Accelerate.framework/Versions/A/Frameworks/vecLib.framework/Versions/A/libBLAS.dylib 
# LAPACK: /Library/Frameworks/R.framework/Versions/4.5-arm64/Resources/lib/libRlapack.dylib;  LAPACK version 3.12.1
# 
# locale:
# [1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8
# 
# time zone: America/New_York
# tzcode source: internal
# 
# attached base packages:
# [1] stats4    stats     graphics  grDevices utils     datasets  methods   base     
# 
# other attached packages:
#  [1] car_3.1-3              carData_3.0-5          ggpmisc_0.6.2          ggpp_0.5.9             openxlsx_4.2.8.1      
#  [6] org.Hs.eg.db_3.21.0    AnnotationDbi_1.70.0   IRanges_2.42.0         S4Vectors_0.46.0       Biobase_2.68.0        
# [11] BiocGenerics_0.54.1    generics_0.1.4         clusterProfiler_4.16.0 pathview_1.48.0        pheatmap_1.0.13       
# [16] GSVA_2.2.1             patchwork_1.3.2        fgsea_1.34.2           DT_0.34.0              edgeR_4.6.3           
# [21] limma_3.64.3           biomaRt_2.64.0         ggfortify_0.4.19       lubridate_1.9.4        forcats_1.0.1         
# [26] stringr_1.6.0          dplyr_1.1.4            purrr_1.2.0            readr_2.1.6            tidyr_1.3.1           
# [31] tibble_3.3.0           ggplot2_4.0.1          tidyverse_2.0.0       
# 
# loaded via a namespace (and not attached):
#   [1] splines_4.5.0               bitops_1.0-9                ggplotify_0.1.3             filelock_1.0.3             
#   [5] R.oo_1.27.1                 graph_1.86.0                XML_3.99-0.20               lifecycle_1.0.4            
#   [9] httr2_1.2.1                 vroom_1.6.6                 MASS_7.3-65                 lattice_0.22-7             
#  [13] crosstalk_1.2.2             magrittr_2.0.4              sass_0.4.10                 rmarkdown_2.30             
#  [17] jquerylib_0.1.4             yaml_2.3.10                 ggtangle_0.0.8              zip_2.3.3                  
#  [21] cowplot_1.2.0               DBI_1.2.3                   RColorBrewer_1.1-3          abind_1.4-8                
#  [25] GenomicRanges_1.60.0        R.utils_2.13.0              RCurl_1.98-1.17             yulab.utils_0.2.1          
#  [29] rappdirs_0.3.3              GenomeInfoDbData_1.2.14     enrichplot_1.28.4           ggrepel_0.9.6              
#  [33] irlba_2.3.5.1               tidytree_0.4.6              MatrixModels_0.5-4          annotate_1.86.1            
#  [37] codetools_0.2-20            DelayedArray_0.34.1         DOSE_4.2.0                  xml2_1.5.0                 
#  [41] tidyselect_1.2.1            aplot_0.2.9                 UCSC.utils_1.4.0            farver_2.1.2               
#  [45] ScaledMatrix_1.16.0         matrixStats_1.5.0           BiocFileCache_2.16.2        jsonlite_2.0.0             
#  [49] Formula_1.2-5               survival_3.8-3              tools_4.5.0                 progress_1.2.3             
#  [53] treeio_1.32.0               Rcpp_1.1.0                  glue_1.8.0                  gridExtra_2.3              
#  [57] SparseArray_1.8.1           xfun_0.54                   qvalue_2.40.0               MatrixGenerics_1.20.0      
#  [61] GenomeInfoDb_1.44.3         HDF5Array_1.36.0            withr_3.0.2                 fastmap_1.2.0              
#  [65] rhdf5filters_1.20.0         SparseM_1.84-2              digest_0.6.39               rsvd_1.0.5                 
#  [69] timechange_0.3.0            R6_2.6.1                    gridGraphics_0.5-1          colorspace_2.1-2           
#  [73] GO.db_3.21.0                dichromat_2.0-0.1           RSQLite_2.4.4               R.methodsS3_1.8.2          
#  [77] h5mread_1.0.1               utf8_1.2.6                  data.table_1.17.8           prettyunits_1.2.0          
#  [81] httr_1.4.7                  htmlwidgets_1.6.4           S4Arrays_1.8.1              pkgconfig_2.0.3            
#  [85] gtable_0.3.6                rsconnect_1.6.2             blob_1.2.4                  S7_0.2.1                   
#  [89] SingleCellExperiment_1.30.1 XVector_0.48.0              htmltools_0.5.8.1           GSEABase_1.70.1            
#  [93] scales_1.4.0                png_0.1-8                   SpatialExperiment_1.18.1    ggfun_0.2.0                
#  [97] knitr_1.50                  rstudioapi_0.17.1           tzdb_0.5.0                  reshape2_1.4.5             
# [101] rjson_0.2.23                nlme_3.1-168                curl_7.0.0                  cachem_1.1.0               
# [105] rhdf5_2.52.1                parallel_4.5.0              pillar_1.11.1               grid_4.5.0                 
# [109] vctrs_0.6.5                 BiocSingular_1.24.0         dbplyr_2.5.1                beachmat_2.24.0            
# [113] xtable_1.8-4                Rgraphviz_2.52.0            evaluate_1.0.5              KEGGgraph_1.68.0           
# [117] magick_2.9.0                cli_3.6.5                   locfit_1.5-9.12             compiler_4.5.0             
# [121] rlang_1.1.6                 crayon_1.5.3                labeling_0.4.3              plyr_1.8.9                 
# [125] fs_1.6.6                    stringi_1.8.7               BiocParallel_1.42.2         Biostrings_2.76.0          
# [129] lazyeval_0.2.2              quantreg_6.1                GOSemSim_2.34.0             Matrix_1.7-4               
# [133] hms_1.1.4                   sparseMatrixStats_1.20.0    bit64_4.6.0-1               Rhdf5lib_1.30.0            
# [137] KEGGREST_1.48.1             statmod_1.5.1               SummarizedExperiment_1.38.1 igraph_2.2.1               
# [141] memoise_2.0.1               bslib_0.9.0                 ggtree_3.16.3               fastmatch_1.1-6            
# [145] bit_4.6.0                   polynom_1.4-1               gson_0.1.0                  ape_5.8-1   
```

